\section{Stream computations}\label{sec:streams}

In this section we introduce formally the notion of a stream as an
infinite sequence of values, and we define computations on streams.
Stream operators (\secref{sec:notation}) are the basic building block of stream computations.  
We employ (\secref{sec:abelian}) restricted types of stream operators: causal operators (which cannot 
``look into the future''), and strict operators (which cannot even ``look into the present'').
Moreover, all our operators are ``synchronous'': they consume and produce data at the same ``rate''. Causal operators can be chained
into complex acyclic computational circuits; cyclic circuits are restricted to using strict operators 
on back-edges.  Finally, we define (\secref{sec:abelianstreams}) two useful stream operators: integration and differentiation.

All the results in this section have been known for decades, 
but we recapitulate them to clarify our model's formal assumptions.

\subsection{Streams and stream operators}\label{sec:notation}

$\N$ is the set of natural numbers, $\B$ is the set of Booleans, and $\Z$ is the set of integers.

\begin{definition}[stream]
Given a set $A$, a \defined{stream} \emph{of values from $A$}, or an \emph{$A$-stream}, is a function $\N \rightarrow A$. 
We denote by $\stream{A} \defn \{ s \,|\, s : \N \to A \}$ the set of all $A$-streams. 
\end{definition}

When $s\in\stream{A}$ and $t\in\N$ we 
write $s[t]$ for the $t$-th element of the stream $s$ instead of the usual $s(t)$
to distinguish it from other function applications.
We think of the index $t\in\N$ as (discrete) time and of $s[t]\in A$ 
as the value of the the stream $s$ ``at time'' $t$.
\ifstreamexamples
For example, the stream of natural numbers $id \in \stream{\N}$ given by $\id[t] = t$ is the sequence of values
$\sv{0 1 2 3 4}$.
\fi

\begin{definition}[stream operator]
A (typed) \defined{stream operator} with $n$ inputs is a function $T:\stream{A_0}\times\cdots\times\stream{A_{n-1}}\to\stream{B}$. 
\end{definition}

In general we will use ``operator'' for functions on streams, and
``function'' for computations on ``scalar'' values.

We are using an extension of the simply-typed lambda calculus to write \dbsp programs;
we will introduce its elements gradually.  However, we find it more readable to
also use signal-processing-like circuit diagrams to depict \dbsp programs.
In a circuit diagram a rectangle represents an operator application (labeled
with the operator name, e.g., $T$), while an arrow is a stream.  

Stream operator \emph{composition} (function composition) is shown as chained circuits.
The composition of a binary operator $T: \stream{A} \times \stream{B} \to \stream{A}$ with the 
unary operator $S: \stream{A} \to \stream{B}$ into the computation 
$\lambda s. T(T(s,S(s)),S(s)) : \stream{A}\to\stream{A}$ 
is:

\begin{center}
\begin{tikzpicture}[auto,>=latex]
  \node[] (input) {$s$};
  \node[] [right of=input] (dummy) {};
  \node[block, below of=dummy, node distance=.7cm] (S1) {$S$};
  \node[block, right of=S1] (T1) {$T$};
  \node[block, right of=T1] (T2) {$T$};
  \node[block, above of=T2, node distance=.7cm] (S2) {$S$};
  \node[right of=T2] (output) {$o$}; 
  \draw[->] (input) -| (S1);
  \draw[->] (input) -| (T1);
  \draw[->] (S1) -- (T1);
  \draw[->] (T1) -- (T2);
  \draw[->] (input) |- (S2);  \draw[->] (T2) -- (output);
  \draw[->] (S2) -- (T2);
\end{tikzpicture}
\end{center}

(Diagrams obscure the \emph{order} of the inputs of an operator; for non-commutative
operators we have to provide more information.)  

\begin{definition}(lifting)
Given a (scalar) function $f: A \to B$,
we define a stream operator $\lift{f} :\stream{A} \to \stream{B}$ 
by \emph{lifting} the function $f$ pointwise in time: $(\lift{f})(s) \defn f \circ s$.
Equivalently, $(\lift{f})(s)[t] \defn f(s[t])$.
This extends to functions of multiple arguments.
\end{definition}

\ifstreamexamples
For example, $(\lift{(\lambda x.(2x))})(id) = \sv{0 2 4 6 8}$.
\fi

\begin{proposition}[distributivity]\label{prop:distributivity}
Lifting distributes over function composition:
$\lift{(f \circ g)} = (\lift{f}) \circ (\lift{g})$.
\end{proposition}
\begin{comment}
\begin{proof}
This is easily proved by using associativity of function composition:
$\forall s . (\lift{(f \circ g)})(s) = (f \circ g) \circ s =
f \circ (g \circ s) = f \circ (\lift{g})(s) = (\lift{f})((\lift{g})(s)) = 
(\lift{f} \circ \lift{g})(s).$
\end{proof}
\end{comment}

We say that two \dbsp programs are \defined{equivalent} if they compute the same
input-output function on streams.
We use the symbol $\cong$ to indicate that two circuits are 
equivalent.  For example, Proposition~\ref{prop:distributivity}
states the following circuit equivalence:

\noindent
\begin{tabular}{m{3.5cm}m{.3cm}m{3.5cm}}
\begin{tikzpicture}[auto,>=latex]
  \node[] (input) {$s$};
  \node[block, right of=input] (g) {$\lift{g}$};
  \node[block, right of=g] (f) {$\lift{f}$};
  \node[right of=f] (output) {$o$};
  \draw[->] (input) -- (g);
  \draw[->] (g) -- (f);
  \draw[->] (f) -- (output);
\end{tikzpicture}
&
$\cong$
&
\begin{tikzpicture}[auto,>=latex]
    \node[] (input) {$s$};
    \node[block, right of=input, node distance=1.5cm] (fg) {$\lift{(f \circ g)}$};
    \node[right of=fg, node distance=1.5cm] (output) {$o$};
    \draw[->] (input) -- (fg);
    \draw[->] (fg) -- (output);
\end{tikzpicture}
\end{tabular}

\subsection{Streams over abelian groups}\label{sec:abelian}

For the rest of the technical development we require the set of values $A$
of a stream $\stream{A}$ to form a commutative group $(A, +, 0, -)$. 
Now we introduce the primitive stream operators that \dbsp uses.

\subsubsection{Delays and time-invariance}\label{sec:delay}

\begin{definition}[Delay]
The \defined{delay operator}\footnote{The name $\zm$
comes from the DSP literature, and is related to the z-transform~\cite{rabiner-book75}.}
produces an output stream
by delaying its input by one step: $\zm_A: \stream{A} \to \stream{A}$:

\begin{tabular}{m{5cm}m{3cm}}
$
\zm_A(s)[t] \defn   \begin{cases}
0        & \text{when}~t=0_A \\
s[t - 1] & \text{when}~t\geq1
\end{cases}
$ &
\begin{tikzpicture}[auto,node distance=1cm,>=latex]
    \node[] (input) {$s$};
    \node[block, right of=input] (z) {$\zm$};
    \node[right of=z] (output) {$o$};
    \draw[->] (input) -- (z);
    \draw[->] (z) -- (output);
\end{tikzpicture}
\end{tabular} 
\end{definition}

We often omit the type parameter $A$, and write just $\zm$.
\ifstreamexamples
For example, $\zm(\id) = \sv{0 0 1 2 3}$.
\fi

\begin{definition}[Time invariance]
A stream operator $S: \stream{A} \to \stream{B}$ is \defined{time-invariant} iff 
$S(\zm_A(s)) = \zm_B(S(s))$ for all $s \in \stream{A}$, or, in other words, iff the
two following circuits are equivalent:

\begin{tabular}{m{3cm}m{.5cm}m{3cm}}
\begin{tikzpicture}[auto,>=latex]
  \node[] (input) {$s$};
  \node[block, right of=input] (S) {$S$};
  \node[block, right of=S] (z) {$\zm$};
  \node[right of=z] (output) {$o$};
  \draw[->] (input) -- (S);
  \draw[->] (S) -- (z);
  \draw[->] (z) -- (output);
\end{tikzpicture}
&
$\cong$
&
\begin{tikzpicture}[auto,>=latex]
  \node[] (input) {$s$};
  \node[block, right of=input] (z) {$\zm$};
  \node[block, right of=z] (S) {$S$};
  \node[right of=S] (output) {$o$};
  \draw[->] (input) -- (z);
  \draw[->] (z) -- (S);
  \draw[->] (S) -- (output);
\end{tikzpicture}
\end{tabular}

\noindent
This definition extends 
naturally to operators with multiple inputs.
\end{definition}

The composition of time-invariant operators of any number of inputs
is time invariant. The delay operator $\zm$ is time-invariant. 
\dbsp only uses time-invariant operators.

\begin{definition}
We say that a function between groups $f: A \to B$ has the \emph{zero-preservation
property} if $f(0_A) = 0_B$.  We write $\zpp{f}$.
\end{definition}

A lifted operator $\lift{f}$ is time-invariant iff $\zpp{f}$.

\subsubsection{Causal and strict operators}\label{sec:causal}

\begin{definition}[Causality]
A stream operator $S:\stream{A}\to\stream{B}$
is \defined{causal} when for all $s,s'\in\stream{A}$,
and all times $t$ we have:
$
(\forall i \leq t~s[i]=s'[i]) ~~\Rightarrow~~ S(s)[t]=S(s')[t].
$
\end{definition}

\noindent
In other words, the output value at time $t$ can only depend on 
input values from times $t' \leq t$.
Operators produced by lifting are causal, and $\zm$ is causal.
All \dbsp operators are causal.  The composition
of causal operators of any number of inputs is causal.

\begin{definition}[Strictness]
A stream operator, $F:\stream{A}\to\stream{B}$ 
is \defined{strictly causal} (abbreviated \textbf{strict})
if  $\forall s,s'\in\stream{A}, \forall t \in \N$ we have:
$(\forall i<t . ~s[i]=s'[i]) ~~\Rightarrow~~ F(s)[t]=F(s')[t].$
\end{definition}

So the $t$-th output of $F(s)$ can depend only on ``past'' values
of the input $s$, between $0$ and $t-1$.  
In particular, $F(s)[0] = 0_B$ is the same for all $s \in \stream{A}$.
Strict operators are causal. Lifted operators in general are \emph{not} strict. 
$\zm$ is strict.  %In \dbsp $\zm$ is the only primitive strict operator.

\begin{proposition}
\label{prop-unique-fix}
For a strict $F: \stream{A} \to \stream{A}$ the equation ~$\alpha=F(\alpha)$~ has a unique
solution $\alpha \in \stream{A}$, denoted by $\fix{\alpha}{F(\alpha)}$. 
\end{proposition}

Thus every strict operator from a set to itself has a unique fixed point.
The simple proof relies on strong induction, showing that $\alpha[t]$ 
depends only on the values of $\alpha$ prior to $t$.  

We show that the following circuit, having a strict ``feedback'' edge $F$,
is a well-defined function on streams:

\begin{center}
\begin{tikzpicture}[>=latex]
    \node[] (input) {$s$};
    \node[block, right of=input] (f) {$T$};
    \node[right of=f] (output) {$\alpha$};
    \node[block, below of=f, node distance=.5cm] (z) {$F$};
    \draw[->] (input) -- (f);
    \draw[->] (f) -- node (mid) {} (output);
    \draw[->] (mid.center) |-  (z);
    \draw[->] (z.west) -- ++(-.3,0) |- ([yshift=1mm]f.south west);
\end{tikzpicture}
\end{center}

\begin{lemma} 
\label{lemma-causal-strict}
If $F: \stream{B} \to \stream{B}$ is strict and $T: \stream{A} \times \stream{B} \to \stream{B}$ is causal, then for a fixed $s$ the operator
$\lambda\alpha.T(s,F(\alpha)): \stream{A} \to \stream{B}$ is strict. 
\end{lemma}

\begin{corollary}\label{feedback-semantics}
\label{cor-loop}
If $F: \stream{B} \to \stream{B}$ is strict and $T: \stream{A} \times \stream{B} \to \stream{B}$ is causal,
the operator $Q(s)=\fix{\alpha}{T(s,F(\alpha))}$ is well-defined and causal. 
If, moreover, $F$ and $T$ are time-invariant then so is $Q$.
\end{corollary}

All stream computations in \dbsp are built from the primitive operators 
we have described: lifted operators and delays (we add two more operators in \secref{sec:nested}).
Circuits composed of such operators can be efficiently implemented 
using Dataflow machines~\cite{lee-ieee95}.

Circuits with feedback are used for two purposes: defining an
integration operator (in the next section), and defining recursive computations (\secref{sec:recursion}).  In turn, the
integration operator will be instrumental in defining incremental
computations (\secref{sec:incremental}).

\subsection{Integration and differentiation}\label{sec:abelianstreams}

Remember that we require the elements of a stream to come from an abelian group $A$.  
Streams themselves form an abelian group:

\begin{proposition}
The structure $(\stream{A},+,0,-)$, obtained by lifting the $+$ and unary $-$ operations from $A$
to $\stream{A}$, is an abelian group.
\end{proposition}

\noindent
Stream addition and negation are causal, time-invariant operators.

\begin{definition}
Given abelian groups $A$ and $B$ we call a stream operator 
$S: \stream{A} \rightarrow \stream{B}$ \defined{linear} if it is a group homomorphism, that is,
$S(a+b)=S(a)+S(b)$ (and therefore $S(0)=0$ and $S(-a)=-S(a)$). 
\end{definition}

Lifting a linear function $f: A \to B$ produces
a stream operator $\lift{f}$ that is linear, time-invariant (LTI).
$\zm$ is LTI.   

\begin{definition}(bilinear)
A function of two arguments $f: A \times B \to C$ with $A, B, C$ groups, is \emph{bilinear} 
if it is linear separately in each argument (i.e., it distributes over addition): 
$\forall a, b, c, d . f(a+b, c) = f(a, c) + f(b, c)$, and $f(a, c+d) = f(a, c) + f(c, d).$
\end{definition}

This definition extends to stream operators.
Lifting a bilinear function $f: A \times B \to C$ produces
a bilinear stream operator $\lift{f}$.  An example bilinear operator over $\stream{\Z}$
is lifted multiplication:
$f: \stream{\N} \times \stream{\N} \to \stream{\N}, f(a, b)[t] = a[t]\cdot b[t]$.  

The composition of (bi)linear operators with linear operators 
is (bi)linear (since homomorphisms compose).

The feedback loop produced with a linear operator is linear:

\begin{proposition}
\label{prop-rec-linear}
Let $S$ be a unary causal LTI operator. The
operator $Q(s)=\fix{\alpha}{S(s+\zm(\alpha))}$ is well-defined and LTI:

\begin{center}
\begin{tikzpicture}[>=latex]
    \node[] (input) {$s$};
    \node[block, shape=circle, right of=input, inner sep=0pt, node distance=.6cm] (plus) {$+$};
    \node[block, right of=plus] (Q) {$S$};
    \node[right of=Q, node distance=1.2cm] (output) {$\alpha$};
    \node[block, below of=Q, node distance=.6cm] (z) {$\zm$};
    \draw[->] (input) -- (plus);
    \draw[->] (plus) -- (Q);
    \draw[->] (Q) -- node (mid) {} (output);
    \draw[->] (mid.center) |-  (z);
    \draw[->] (z) -| (plus);
\end{tikzpicture}
\end{center}
\end{proposition}

\begin{definition}[Differentiation]
The \defined{differentiation operator} $\D_{\stream{A}} : \stream{A} \to \stream{A}$ is defined by:
$\D(s) \defn s - \zm(s)$.
\end{definition}
We generally omit the type, and write just $\D$ when the type can be inferred from the context.
The value of $\D(s)$ at time $t$ is the difference
between  the current (time $t$) value of $s$ and the previous (time $t-1$) value of $s$.
\ifstreamexamples
As an example, $\D(\id) = \sv{0 1 1 1 1}$.
\fi

If $s$ is a stream, then $\D(s)$ is the \emph{stream of changes} of $s$. 

\begin{proposition}
\label{prop-diff-properties}
Differentiation $\D$ is causal and LTI.
\end{proposition}

\begin{tabular}{cc}
\begin{tikzpicture}[auto,>=latex,node distance=1cm]
    \node[] (input) {$s$};
    \node[block, shape=circle, right of=input, inner sep=0pt,node distance=2cm] (plus) {$+$};
    \node[right of=plus] (output) {$\D(s)$};
    \draw[draw,->] (input) -- node (i) {} (plus);
    \node[block, below of=i, node distance=.7cm] (z) {$\zm$};
    \node[block, shape=circle, right of=z, inner sep=1pt] (minus) {$-$};
    \draw[->] (plus) -- (output);
    \draw[->] (i) -- (z);
    \draw[->] (z) -- (minus);
    \draw[->] (minus) -- (plus);
\end{tikzpicture} &
\begin{tikzpicture}[auto,>=latex]
    \node[] (input) {$s$};
    \node[block, shape=circle, right of=input, inner sep=0pt] (plus) {$+$};
    \node[right of=plus, node distance=1.5cm] (output) {$\I(s)$};
    \node[block, below of=plus, node distance=.7cm] (z) {$z^{-1}$};
    \draw[draw,->] (input) -- (plus);
    \draw[->] (plus) -- node (o) {} (output);
    \draw[->] (o) |- (z);
    \draw[->] (z) -- (plus);
\end{tikzpicture} \\
Differentiation & Integration
\end{tabular}

The integration operator ``reconstitutes'' a stream from its changes:

\begin{definition}[Integration]
The \defined{integration operator}  $\I_{\stream{A}} : \stream{A} \to \stream{A}$ 
is defined by $\I(s) \defn \lambda s . \fix{\alpha}{(s + \zm(\alpha))}$.
\end{definition}

\noindent
We also generally omit the type, and write just $\I$.
This is the construction from Proposition~\ref{prop-rec-linear} 
using the identity function for $S$.

\begin{proposition}
$\I(s)$ is the discrete (indefinite) integral applied to the stream $s$: 
$\I(s)[t] = \sum_{i \leq t} s[i]$.
\end{proposition}
\ifstreamexamples
As an example, $\I(\id) = \sv{0 1 3 6 10}$.
\fi

\begin{proposition}
\label{prop-integ-properties}
$\I$ is causal and LTI.
\end{proposition}

\begin{theorem}[Inversion]
\label{inverses}
Integration and differentiation are inverses of each other:
$\forall s . \I(\D(s)) = \D(\I(s)) = s$.
\end{theorem}

\noindent
\begin{tabular}{m{2.5cm}m{.3cm}m{1cm}m{.3cm}m{2.5cm}}
\begin{tikzpicture}[auto,>=latex, node distance=.75cm]
    \node[] (input) {$s$};
    \node[block, right of=input] (I) {$\I$};
    \node[block, right of=I] (D) {$\D$};
    \node[right of=D] (output) {$o$};
    \draw[->] (input) -- (I);
    \draw[->] (I) -- (D);
    \draw[->] (D) -- (output);
\end{tikzpicture}
     &  
     $\cong$
     &
     \hspace{-2ex}
\begin{tikzpicture}[auto,>=latex, node distance=.75cm]
    \node[] (input) {$s$};
    \node[right of=input] (output) {$o$};
    \draw[->] (input) -- (output);
\end{tikzpicture}
     &
     $\cong$
     &
\begin{tikzpicture}[auto,>=latex, node distance=.75cm]
    \node[] (input) {$s$};
    \node[block, right of=input] (D) {$\D$};
    \node[block, right of=D] (I) {$\I$};
    \node[right of=I] (output) {$o$};
    \draw[->] (input) -- (D);
    \draw[->] (D) -- (I);
    \draw[->] (I) -- (output);
\end{tikzpicture}
\end{tabular}

\section{Incremental computation}\label{sec:incremental}
 
\begin{definition}
Given a unary stream operator $Q: \stream{A} \to \stream{B}$ we define the 
\defined{incremental version} of $Q$ as $\inc{Q} \defn \D \circ Q \circ \I$.
$\inc{Q}$ has the same ``type'' as $Q$: $\inc{Q}: \stream{A} \to \stream{B}$.
For an operator with multiple inputs we define 
the incremental version by applying $\I$ to each input independently:
e.g., if $T: \stream{A} \times \stream{B} \rightarrow \stream{C}$ then
$\inc{T}(a, b) \defn \D (T(\I(a), \I(b)))$.
\end{definition}

The following diagram illustrates the intuition behind this definition: 
\begin{tikzpicture}[auto,>=latex]
    \node[] (input) {$\Delta s$};
    \node[block, right of=input] (I) {$\I$};
    \node[block, right of=I] (Q) {$Q$};
    \node[block, right of=Q] (D) {$\D$};
    \node[right of=D] (output) {$\Delta o$};
    \draw[->] (input) -- (I);
    \draw[->] (I) -- node (s) {$s$} (Q);
    \draw[->] (Q) -- node (o) {$o$} (D);
    \draw[->] (D) -- (output);
\end{tikzpicture}

If $Q(s) = o$ is a computation, then $\inc{Q}$ performs 
the ``same'' computation as $Q$,
but between streams of changes $\Delta s$ and $\Delta o$.
This is the diagram from the
introduction, substituting $\Delta s$ for the transaction stream $T$, 
and $o$ for the stream of view versions $V$.

Notice that our definition of incremental computation is meaningful only for \emph{streaming}
computations; this is in contrast to classic definitions, e.g.~\cite{gupta-idb95} which
consider only one change.  Generalizing the definition to operate on streams gives us
additional power, especially when operating with recursive queries.

The following proposition is one of our central results.

\begin{proposition}(Properties of the incremental version):
For computations of appropriate types, the following hold:
\label{prop-inc-properties}
\begin{description}[nosep, leftmargin=\parindent]
\item[inversion:] $Q\mapsto\inc{Q}$ is bijective; its inverse is $Q\mapsto \I\circ Q\circ\D$.
\item[invariance:] $\inc{+} = +, \inc{(\zm)} = \zm, \inc{-} = -, \inc{\I}=\I, \inc{\D}=\D$
\item[push/pull:] \label{prop-part-commutation}
    $Q \circ \I = \I \circ \inc{Q}$; $\D\circ Q = \inc{Q}\circ\D$
\item[chain:] $\inc{(Q_1\circ Q_2)} = \inc{Q_1}\circ\inc{Q_2}$ (This generalizes to operators with multiple inputs.)
\item[add:] $\inc{(Q_1 + Q_2)} = \inc{Q_1} + \inc{Q_2}$
\item[cycle:] $\inc{(\lambda s. \fix{\alpha}{T(s,\zm(\alpha)}))} = \lambda s. \fix{\alpha}{\inc{T}(s,\zm(\alpha)})$
\end{description}
\end{proposition}

The proof of these properties relies on elementary algebraic manipulations.
Despite their simplicity, they are very useful.
For example, the \defined{chain rule}  states that the following two circuits are equivalent:

\noindent
\begin{tabular}{m{4cm}m{.2cm}m{2.5cm}}
\begin{tikzpicture}[auto,>=latex,node distance=.8cm]
  \node[] (input) {$i$};
  \node[block, right of=input] (I) {$\I$};
  \node[block, right of=I] (Q1) {$Q_1$};
  \node[block, right of=Q1] (Q2) {$Q_2$};
  \node[block, right of=Q2] (D) {$\D$};
  \node[right of=D] (output)  {$o$};
  \draw[->] (input) -- (I);
  \draw[->] (I) -- (Q1);
  \draw[->] (Q1) -- (Q2);
  \draw[->] (Q2) -- (D);
  \draw[->] (D) -- (output);
\end{tikzpicture} &
$\cong$ &
\begin{tikzpicture}[>=latex]
  \node[] (input) {$i$};
  \node[block, right of=input] (Q1) {$\inc{Q_1}$};
  \node[block, right of=Q1] (Q2) {$\inc{Q_2}$};
  \node[right of=Q2] (output)  {$o$};
  \draw[->] (input) -- (Q1);
  \draw[->] (Q1) -- (Q2);
  \draw[->] (Q2) -- (output);
\end{tikzpicture}
\end{tabular}

\noindent In other words, \textbf{to incrementalize a composite query you can incrementalize
each sub-query independently}.  This gives us a simple deterministic recipe  
for computing the incremental version of an arbitrarily complex query.

We illustrate by giving the proof of the 
chain rule, which is trivial, and is based on 
function composition associativity:
$$
\begin{aligned}
\inc{(Q_1 \circ Q_2)} &= \D \circ Q_1 \circ Q_2 \circ \I \\
 &= \D \circ Q_1 \circ (\I \circ \D) \circ Q_2 \circ \I \\
 &= (\D \circ Q_1 \circ \I) \circ (\D \circ Q_2 \circ \I) \\
 &= \inc{(Q_1)} \circ \inc{(Q_2)}.
\end{aligned}
$$

The \defined{cycle rule} states that the following circuits are equivalent:

\noindent
\begin{tabular}{m{4.2cm}m{.2cm}m{3cm}}
\begin{tikzpicture}[>=latex]
    \node[] (input) {$s$};
    \node[block, right of=input] (I) {$\I$};
    \node[block, right of=I] (f) {$T$};
    \node[block, right of=f] (D) {$\D$};
    \node[right of=D] (output) {$o$};
    \node[block, below of=f, node distance=.7cm] (z) {$\zm$};
    \draw[->] (input) -- (I);
    \draw[->] (I) -- (f);
    \draw[->] (f) -- node (mid) {} (D);
    \draw[->] (mid.center) |-  (z);
    \draw[->] (z.west) -- ++(-.3,0) |- ([yshift=1mm]f.south west);
    \draw[->] (D) -- (output);
\end{tikzpicture} & $\cong$ &
\begin{tikzpicture}[>=latex]
    \node[] (input) {$s$};
    \node[block, right of=input] (f) {$\inc{T}$};
    \node[right of=f, node distance=1.5cm] (output) {$o$};
    \node[block, below of=f, node distance=.7cm] (z) {$\zm$};
    \draw[->] (input) -- (f);
    \draw[->] (f) -- node (mid) {} (output);
    \draw[->] (mid.center) |-  (z);
    \draw[->] (z.west) -- ++(-.3,0) |- ([yshift=1mm]f.south west);
\end{tikzpicture}
\end{tabular}

The incremental version of a feedback loop around a query
is just the feedback loop with the incremental query.  The significance
of this result will be apparent when we implement recursive queries.

To execute incremental queries efficiently, we want to compute directly 
on streams of changes without integrating them. The invariance property above shows
that stream operators $+$, $-$, and $\zm$ are identical to their incremental versions, 
thus $\I$ and $\D$ can be omitted for them: $\inc{Q} = \I\circ Q\circ\D = Q$.
The following theorems generalize this to linear and bi-linear operators:

\begin{theorem}[Linear]\label{linear}
For an LTI operator $Q$ we have $\inc{Q}=Q$.
\end{theorem}

\begin{theorem}[Bilinear]\label{bilinear}
For a bilinear time-invariant operator $\times$ we have
$\inc{(a \times b)} ~=~ a \times b ~+~ \zm(\I(a)) \times b ~+~ a \times \zm(\I(b))$.
\end{theorem}

By rewriting this statement using $\Delta a$ for the stream of changes to $a$ we
get the familiar formula for incremental equi-joins:
$\Delta(a\times b) =\Delta a \times \Delta b + a\times(\Delta b) + (\Delta a)\times b$.

This should not be surprising because equi-joins are bilinear,
as we discuss in the next section.  
